A Shared Subspace Model for Multi-Group Covariance Estimation

### 2016-07-09 paper/mscov.tex
Prettifyed figures.  Fixed typos.

\title{Shared Subspace Models for Multi-Group Covariance Estimation
%
  \protect\thanks{Alexander M. Franks is a Moore/Sloan Data Science
    and WRF Innovation in Data Science Postdoctoral Fellow and a
    graduate of the Department of Statistics at Harvard University
    (\href{mailto:amfranks@uw.edu}{amfranks@uw.edu}).  Peter D. Hoff
    is a Professor in the Departments of Statistics, Biostatistics and
    the Center for Statistics and the Social Sciences at the
    University of Washington
    (\href{mailto:pdhoff@uw.edu}{pdfhoff@uw.edu}).
%
This work was partially supported 
 by the Washington Research Foundation Fund for Innovation in Data-Intensive Discovery and the Moore/Sloan Data Science Environments Project at the University of Washington.
%
 The authors are grateful to Dr.\ Daniel Promislow (Department of
 Pathology, University of Washington) and Dr.\ Jessica Hoffman
 (University of Alabama at Birmingham) for sharing data and ideas
 which contributed to framing of this paper.}}
\author{Alexander Franks and Peter Hoff} \date{\today}
\maketitle

\begin{abstract}

@@

  We develop a model-based method for evaluating heterogeneity among
  several $p\times p$ covariance matrices in the big $p$, small $n$ setting.
  This is done by assuming a spiked covariance model for each group
  and sharing information about the space spanned by the group-level
  eigenvectors.  We use an empirical Bayes method to identify a low-dimensional
  subspace which explains variation across all groups and use an MCMC
  algorithm to estimate the posterior uncertainty of eigenvectors and
  eigenvalues on this subspace.  The implementation and utility of our
  model is illustrated with analyses of high-dimensional multivariate
  gene expression and metabolomics data.
\vfill
\noindent\textbf{Keywords:} covariance estimation; spiked covariance
model; Stiefel manifold; Grassmann manifold; big p, small n;
high-dimensional data; empirical Bayes.
\end{abstract}

@@

Multivariate data is often partitioned into
groups, each of which represent samples from populations with
distinct, but possibly related, distributions.  Although historically the
primary focus has been on identifying mean-level differences between
populations, there has been a growing need to identify differences in
population covariances as well.  For instance, in case-control studies, mean level effects may be small relative to subject
variability; distributional differences between groups may still be
evident as differences in the covariances between features.  Even when
mean level differences are detectable, better estimates of the
covariability of features across groups may lead to an improved
understanding of the mechanisms underlying these apparent mean level
differences.  Further, accurate covariance estimation is an essential
part of many prediction tasks (e.g. quadratic discriminant analysis).
Thus, evaluating heterogeneity between covariance
matrices can be an important complement to more traditional analyses
for estimating differences in means across
groups.

To address this need, we develop a novel method for
group-covariance estimation.  In our method, we exploit the fact that
in many natural systems, high dimensional data is often very
structured and thus can be best understood on a lower dimensional
subspace. For example, with gene expression data, the effective
dimensionality is thought to scale with the number of gene regulatory
modules, not the number of genes themselves \citep{Heimberg2016}.  As
such, differences in gene expression across groups should be expressed
in terms of differences between these regulatory modules rather than
strict differences between expression levels.  Such differences can be
examined on a subspace that reflects the correlations resulting from
these modules.  In contrast to most existing approaches for group
covariance estimation, our approach is to directly infer such
subspaces from groups of related data.

@@

Another popular method for modeling relationships between
high-dimensional multivariate data is partial least squares regression
(PLS) \citep{Wold2001}. This approach, which is a special case of a bilinear factor
model, involves projecting the data onto a lower dimensional space which
maximizes the similarity of the two groups.  This technique does not
require the data from each group to share the same feature set.  A
common variant for prediction, partial least squares discriminant
analysis (PLS-DA) is especially common in chemometrics and
bioinformatics \citep{Barker2003} .  Although closely related to the
approaches we will consider here, the primarily goal of of PLS-based
models is to create regression or discrimination models, not to
explicitly infer covariance matrices from multiple groups of data.
Nevertheless, the basic idea that data can often be well represented
on a low dimensional space is an appealing one that we leverage.

@@

In Section \ref{sec:simulation} we investigate the behavior of this
class of models in simulation and demonstrate how the shared subspace
assumption is widely applicable, even when there is little similarity
in the covariance matrices across groups.  In particular, shared
subspace models include those in which there is no information shared
across groups because in the worst case we can assume the subspace
spanned by the group-level eigenvectors is simply the full
$p$-dimensional space.  In Section \ref{sec:app} we demonstrate the
utility of a shared subspace model in an analysis of gene expression
data from juvenile leukemia patients .  Despite the large feature size
($p > 3000$) and small sample size ($n < 100$ per group), we identify
interpretable similarities and differences in gene covariances on a
low dimensional subspace.  Finally, we conclude with a brief
exploration of heterogeneity in metabolomic data from a study on aging
in the fly species \textit{Drosophila melanogaster}.

@@

Suppose $S$ has a possibly degenerate Wishart$(\Sigma,n$)
distribution with density given by
\begin{equation} 
p(S | \Sigma, n) \propto l(\Sigma: S) =  |\Sigma|^{-n/2} \text{etr}( - \Sigma^{-1}  S/2 ) ,  
\label{eqn:lik}
\end{equation}
%
\noindent where $\Sigma \in \mathcal S_p^+$ and $n$ may be less than $p$.  Such
a likelihood results from $S$ being, for example, a residual sum
of squares matrix from a multivariate regression analysis. In this
case, $n$ is the number of independent observations minus the rank of
the design matrix.  The spiked principle components model (spiked PCA)
studied by \citet{Johnstone2001} and others assumes that
\begin{equation} 
\Sigma = \sigma^2 ( U  \Lambda  U^T  + I )
\label{eqn:spiked}
\end{equation}
%
\noindent where for $r \ll p$, $\Lambda$ is an $r\times r$ diagonal
matrix and $U \in \mathcal V_{p,r}$.  $\mathcal V_{p,r}$ is the
Stiefel manifold, which consists of the set of $p \times r$
orthonormal matrices in $\mathbb{R}^p$, so that $U^TU =
I$.  The spiked covariance formulation is appealing because it explicitly
partitions the covariance matrix into a tractable low rank ``signal''
and isotropic ``noise''.

@@

Classical results for parametric models (e.g., \citet{Schwartz1965})
imply that asymptotically in $n$ for fixed $p$, an estimator will be
consistent for a spiked population covariance as long as the assumed
number of spikes (eigenvalues larger than $\sigma^2$) is greater than
or equal to the true number.  However, when $p$ is large relative to
$n$, as is the case for the examples considered here, things are more
difficult.  Under the spiked covariance model, it has been shown that
if $p/n \rightarrow \alpha >0$ as $n\rightarrow \infty$, the $k$th
largest eigenvalue of $S/(n\sigma^2)$ will converge to an upwardly
biased version of $\lambda_{k}+1$, if $\lambda_k$ is greater than
$\sqrt{\alpha}$ \citep{Baik2006, Paul2007} .  This has led several
authors to suggest estimating $\Sigma$ via shrinkage of the
eigenvalues of the sample covariance matrix. In particular, in the
setting that $\sigma^2$ is known, \citet{Donoho2013} propose
estimating all eigenvalues whose sample estimates are smaller than
$\sigma^2(1+\sqrt{\alpha})^2$ by $\sigma^2$, and shrinking the larger
eigenvalues in a way that depends on the particular loss function
being used.  These shrinkage functions are shown to be asymptotically
optimal in the $p/n\rightarrow \alpha$ setting.

Note that covariance estimators of this form are are equivariant with
respect to rotations and scale changes.  The situation should be
different, however, when we are interested in estimating multiple
covariance matrices from distinct, but related groups.  Here,
group-level equivariance to rotations is an unreasonable assumption;
both eigenvalue \emph{and} eigenvector shrinkage can play an important
role in improving covariance estimates.  Consider multi-group
covariance estimation based on $K$ matrices, $Y_1, ..., Y_K$, where
$Y_k$ is assumed to be an $n_k$ by $p$ matrix of mean-zero normal
data, typically with $n_k \ll p$.  Then, $Y_k^TY_k = S_k$ has a
(degenerate) Wishart distribution as in Equation \ref{eqn:lik}.  To
improve estimation, we seek estimators of each covariance matrix,
$\hat{\Sigma}_k$, that may depend on data from all groups.  To this
end, we extend the spiked covariance model to grouped data, by
assuming that the anisotropic variability from each group occurs on a
common low dimensional subspace.  Specifically, we posit that the
covariance matrix for each group can be written as
\begin{equation}
\Sigma_k = V\Psi_kV^T + \sigma^2_kI,
\label{eqn:sspsi}
\end{equation}

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/3dplot}
        \caption{Projection in $\mathcal{R}^3$}
        \label{fig:3dplot}
    \end{subfigure}
\quad
 %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter}
        \caption{$Y_kV$}
        \label{fig:2dscatter}
    \end{subfigure}
\quad 
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter-orth}
        \caption{$Y_kV_{\perp}$}
        \label{fig:2dscatterorth}
    \end{subfigure}
    \caption{ Two groups of four-dimensional data (red and blue)
      projected into different subspaces.  a) To visualize $Y_k$ we
      can project the data into $R^3$.  In this illustration, the
      distributional differences between the groups are confined to a
      two-dimensional shared subspace ($VV^T$, grey plane).  b) The
      data projected onto the two-dimensional shared subspace,
      $Y_kV$, have covariances $\Psi_k$ that differ between
      groups. c) The orthogonal projection, $Y_kV_{\perp}$
      has isotropic covariance, $\sigma_k^2I$, for all groups.  }
\label{fig:shared}
\end{figure}
% For multi-group covariance estimation, we assume that we have $K$
% matrices, $Y_1, ... Y_K$, where $Y_k$ consists of $n_k$ independent
% rows of mean-zero normal data of dimension $p$, typically with
% $n_k \ll p$.

%
\noindent where, the columns of $V \in \mathcal{V}_{p, s}$, with
$s \ll p$, determine a subspace shared by all groups.  Throughout this
paper we will denote to the shared subspace as
$VV^T \in \mathcal G_{p,s}$, where $\mathcal G_{p,s}$, known as the
Grassmann manifold, consists of all $s$-dimensional linear subspaces
of $\mathbb{R}^p$ \citep{Chikuse2012}.  Although $V$ is only
identifiable up to right rotations, the matrix $VV^T$, which defines
the plane of variation shared by all groups, is identifiable.  Later,
to emphasize the connection to the spiked PCA model
(\ref{eqn:spiked}), we will write $\Psi_k$ in terms of its
eigendecomposition, $\Psi_k = O_k\Lambda_kO_k$, where $O_k$ are
eigenvectors and $\Lambda_k$ are the eigenvalues of $\Psi_k$ (see
Section \ref{sec:bayes}).

@@

In this section we outline an empirical Bayesian approach for
estimating a low-dimensional shared subspace and the covariance
matrices of the data projected onto this space. As we highlight in
Section \ref{sec:simulation}, if the spiked covariance model holds for
each group individually, then the shared subspace assumption also
holds because at worst we can take the shared subspace to be the full
$p$-dimensional space. In practice, we can usually identify a shared
subspace of dimension $s \ll p$ that preserves most of the variation
in the data.  Our primary objective is to identify the ``best'' shared
subspace of fixed dimension $s < p$.  In Section \ref{sec:em} we
describe an expectation maximization algorithm for estimating the
maximum marginal likelihood of the shared subspace, $VV^T$.  This
approach is computationally tractable for high-dimensional datasets.
Given an inferred subspace, we then seek estimators for the covariance
matrices of the data projected onto this space.  Because seemingly
large differences in the point estimates of covariance matrices across
groups may not actually reflect statistically significant differences,
in Section \ref{sec:bayes} we also describe a Gibbs sampler that can
be used to generate estimates of the projected covariance matrices,
$\Psi_k$, and and their associated uncertainty.  Later, we discuss
strategies for infering an appropriate value for $s$ and explore how
shared subspace models can be used for exploratory data analysis by
visualizing covariance heterogeneity on two or three dimensional
subspaces.

@@

\begin{align}
\nonumber p(S_1, ... S_k | \Sigma_k,n_k) &\propto \prod_{k=1}^K |\Sigma_k|^{-n_k/2}\etr(-\Sigma_k^{-1}S_k/2)  \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-(\sigma_k^2(V\Psi_kV^T +
  I))^{-1}S_k/2) \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-\left[V(\Psi_k +
  I)^{-1}/\sigma_k^2 V^T + (I-VV^T)/\sigma^2_k\right]S_k/2)
  \\
&\propto \prod_{k=1}^K  (\sigma_k^2)^{-n_k(p-s)/2}|M_k|^{-n_k/2}\etr(-\left[VM_k^{-1}V^T + \frac{1}{\sigma^2_k} (I-VV^T)\right]S_k/2) ,
\end{align}
%
\noindent where we define $M_k = \sigma_k^2(\Psi_k + I)$.  The log-likelihood in
$V$ (up to an additive constant) is
%
\begin{align}
\nonumber l(V) &= \sum_k \tr\left(-VM_k^{-1}V^T +
       VV^T/\sigma^2_k\right)S_k/2)\\
&= \sum_k \frac{1}{2}\tr\left(-M_k^{-1}V^T S_kV\right) + \frac{1}{2\sigma_k^2}tr\left(VV^T S_k\right).
\end{align}

@@

We maximize the marginal likelihood of $V$ with an EM algorithm, where
$M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ as the considered the
``missing'' parameters to be imputed.  We assume independent Jeffreys'
prior distributions for both $\sigma_k^2$ and $M_k$.  Jeffreys' prior
for these quantities corresponds to
$p(\sigma_k^2) \propto 1/\sigma_k^2$ and
$p(M_k) \propto |M_k|^{-(s+1)/2}$.  From the likelihood it can easily
be shown that the conditional posterior for $M_k$ is
%
$$p(M_k | V) \propto |M_k|^{(n_k + s + 1)/2}\etr(-1/2(M_k^{-1}V^TS_kV)) $$
%
\noindent which is an inverse-Wishart($V^TS_kV$, $n_k$) distribution.  The
conditional posterior distribution of $\sigma_k^2$ is simply
%
$$p\left(\sigma^2 | V\right) \propto (\sigma_k^2)^{-n_k(p-s)/2-1}\etr\left(-\frac{1}{2\sigma^2_k} (I-VV^T)S_k\right)  $$
%
\noindent which is an inverse-gamma($n_k(p-s)/2$,
$\frac{1}{2}\text{tr}[(I-VV^T)S_k]$) distribution.  Based on these
results, we complete the following steps for each iteration, $t$, of
the EM algorithm until convergence:

@@

For the second step (``M-step''), we use a numerical optimization algorithm based on
the Cayley transform to preserve the orthogonality constraints in $V$
\citep{Wen2013}.  Importantly, the complexity of this algorithm is
dominated by the dimension of the shared subspace not the number of
features.  Thus, our approach is computationally efficient for
relatively small values of $s$, even when $p$ is large.

@@

If $V$ is a basis for a shared subspace, then for each group, most
of the non-isotropic variation in $Y_k$ should be preserved when
projecting the data to this space.  To characterize the extent to
which this is true for different groups, we propose a simple estimator
for the proportion of ``signal'' variance that lies on a given
subspace.  Specifically, we use the following statistic for the ratio
of the sum of the first $s$ eigenvalues of
$V^T \Sigma_k V$ to the sum of the first $s$ eigenvalues
of $\Sigma_k$:
%
\begin{equation}
\gamma(Y_k: V, \sigma_k^2) = \frac{||Y_kV||_F/n_k}{\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}
  ||Y_k\widetilde{V}||_F/n_k - \sigma_k^2ps/n_k}
\label{eqn:ratio}
\end{equation}
%
\noindent where $||\cdot||_F$ is the Frobenius norm. Our motivation for
this statistic is the fact that if
$\Sigma_k = V\Psi_kV^T + \sigma_k^2I$ then
$\gamma(Y_k: V, \sigma_k^2) \approx 1$.  We establish
this in an asymptotic regime where $s$ is fixed, $p$,
$n_k \rightarrow \infty$ and $p/n_k = \alpha_k$ constant.

@@

As a consequence, the proposed statistic will be close to one for all
groups when $VV^T$ is a shared subspace for the data and smaller if
not.  The metric provides a useful indicator for which groups can be
reasonably compared on a given subspace and which groups cannot.  In
practice, we estimate a shared subspace, $\hat{V}$, and the isotropic
variances, $\hat{\sigma}_k^2$, using EM and compute the plug-in
estimate $\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2)$.  When this
statistic is small for some groups, it may suggest that the rank, $s$,
of the inferred subspace needs to be larger to capture the variation
in all groups. We investigate this in Section \ref{sec:simulation}, by
computing the goodness of fit statistic for inferred subspaces of
different dimensions on a single dataset. In Section \ref{sec:app}, we
compute the estimates for subspaces inferred with real biological
data.

@@

The EM algorithm presented in the previous section yields point
estimates for $VV^T$ and $\Psi_k$, but does not lead to natural
uncertainty quantification for these estimates.  In this section, we
assume that the subspace $VV^T$ is fixed and known and demonstrate how
we can estimate the posterior distribution for $\Psi_k$.   Note that when
the subspace is known, the posterior distribution of $\Sigma_k$ is
conditionally independent from the other groups, so that we can
independently estimate the conditional posterior distributions for each
group.

>>>>

\title{Shared Subspace Models for Multi-Group Covariance Estimation
%
  \protect\thanks{Alexander M. Franks is a Moore/Sloan Data Science
    and WRF Innovation in Data Science Postdoctoral Fellow
    (\href{mailto:amfranks@uw.edu}{amfranks@uw.edu}).  Peter D. Hoff
    is a Professor at the Departments of Statistics and Biostatistics
    at the University of Washington
    (\href{mailto:pdhoff@uw.edu}{pdfhoff@uw.edu}).
%
This work was partially supported 
 by the Washington Research Foundation Fund for Innovation in
 Data-Intensive Discovery, the Moore/Sloan Data Science Environments
 Project at the University of Washington and NSF grant DMS-1505136.
%
 The authors are grateful to Dr.\ Daniel Promislow (Department of
 Pathology, University of Washington), Dr.\ Jessica Hoffman
 (University of Alabama at Birmingham) and Dr. Julie Jossie (INRIA)
 for sharing data and ideas which contributed to framing of this
 paper.}}  \author{Alexander Franks and Peter Hoff} \date{\today}
\maketitle 
\begin{abstract}

@@

  We develop a model-based method for evaluating heterogeneity among
  several $p\times p$ covariance matrices in the big $p$, small $n$ setting.
  This is done by assuming a spiked covariance model for each group
  and sharing information about the space spanned by the group-level
  eigenvectors.  We use an empirical Bayes method to identify a low-dimensional
  subspace which explains variation across all groups and use an MCMC
  algorithm to estimate the posterior uncertainty of eigenvectors and
  eigenvalues on this subspace.  The implementation and utility of our
  model is illustrated with analyses of high-dimensional multivariate
  gene expression and metabolomics data.
\\
\\
\noindent\textbf{Keywords:} covariance estimation; spiked covariance
model; Stiefel manifold; Grassmann manifold; big $p$, small $n$;
high-dimensional data; empirical Bayes.
\vfill
\end{abstract}

@@

Multivariate data is often partitioned into
groups, each of which represent samples from populations with
distinct but possibly related distributions.  Although historically the
primary focus has been on identifying mean-level differences between
populations, there has been a growing need to identify differences in
population covariances as well.  For instance, in case-control studies, mean-level effects may be small relative to subject
variability; distributional differences between groups may still be
evident as differences in the covariances between features.  Even when
mean-level differences are detectable, better estimates of the
covariability of features across groups may lead to an improved
understanding of the mechanisms underlying these apparent mean-level
differences.  Further, accurate covariance estimation is an essential
part of many prediction tasks (e.g. quadratic discriminant analysis).
Thus, evaluating heterogeneity between covariance
matrices can be an important complement to more traditional analyses
for estimating differences in means across
groups.

To address this need, we develop a novel method for
multi-group covariance estimation.  Our method exploits the fact that
in many natural systems, high dimensional data is often very
structured and thus can be best understood on a lower dimensional
subspace. For example, with gene expression data, the effective
dimensionality is thought to scale with the number of gene regulatory
modules, not the number of genes themselves \citep{Heimberg2016}.  As
such, differences in gene expression across groups should be expressed
in terms of differences between these regulatory modules rather than
strict differences between expression levels.  Such differences can be
examined on a subspace that reflects the correlations resulting from
these modules.  In contrast to most existing approaches for group
covariance estimation, our approach is to directly infer such
subspaces from groups of related data.

@@

Another popular method for modeling relationships between
high-dimensional multivariate data is partial least squares regression
(PLS) \citep{Wold2001}. This approach, which is a special case of a bilinear factor
model, involves projecting the data onto a lower dimensional space which
maximizes the similarity of the two groups.  This technique does not
require the data from each group to share the same feature set.  A
common variant for prediction, partial least squares discriminant
analysis (PLS-DA) is especially common in chemometrics and
bioinformatics \citep{Barker2003}.  Although closely related to the
approaches we will consider here, the primarily goal of PLS-based
models is to create regression or discrimination models, not to
explicitly infer covariance matrices from multiple groups of data.
Nevertheless, the basic idea that data can often be well represented
on a low dimensional space is an appealing one that we leverage.

@@

In Section \ref{sec:simulation} we investigate the behavior of this
class of models in simulation and demonstrate how the shared subspace
assumption is widely applicable, even when there is little similarity
in the covariance matrices across groups.  In particular, independent
covariance estimation is equivalent to shared subspace
estimation with a sufficiently large shared subspace.  In Section
\ref{sec:app} we demonstrate the utility of a shared subspace model in
an analysis of gene expression data from juvenile leukemia patients .
Despite the large feature size ($p > 3000$) and small sample size
($n < 100$ per group), we identify interpretable similarities and
differences in gene covariances on a low dimensional subspace.
Finally, we conclude with a brief exploration of heterogeneity in
metabolomic data from a study on aging in the fly species
\textit{Drosophila melanogaster}.

@@

Suppose a random matrix $S$ has a possibly degenerate Wishart$(\Sigma,n$)
distribution with density given by
\begin{equation} 
p(S | \Sigma, n) \propto l(\Sigma: S) =  |\Sigma|^{-n/2} \text{etr}( - \Sigma^{-1}  S/2 ) ,  
\label{eqn:lik}
\end{equation}
%
\noindent where $\Sigma \in \mathcal S_p^+$ and $n$ may be less than $p$.  Such
a likelihood results from $S$ being, for example, a residual sum
of squares matrix from a multivariate regression analysis. In this
case, $n$ is the number of independent observations minus the rank of
the design matrix.  The spiked principal components model (spiked PCA)
studied by \citet{Johnstone2001} and others assumes that
\begin{equation} 
\Sigma = \sigma^2 ( U  \Lambda  U^T  + I )
\label{eqn:spiked}
\end{equation}
%
\noindent where for $r \ll p$, $\Lambda$ is an $r\times r$ diagonal
matrix and $U \in \mathcal V_{p,r}$, where $\mathcal V_{p,r}$ is the
Stiefel manifold consisting of all $p \times r$ orthonormal matrices in $\mathbb{R}^p$, so that
$U^TU = I_r$.  The spiked covariance formulation is appealing because it
explicitly partitions the covariance matrix into a tractable low rank
``signal'' and isotropic ``noise''.

@@

Classical results for parametric models (e.g., \citet{Schwartz1965})
imply that asymptotically in $n$ for fixed $p$, an estimator will be
consistent for a spiked population covariance as long as the assumed
number of spikes (eigenvalues larger than $\sigma^2$) is greater than
or equal to the true number.  However, when $p$ is large relative to
$n$, as is the case for the examples considered here, things are more
difficult.  Under the spiked covariance model, it has been shown that
if $p/n \rightarrow \alpha >0$ as $n\rightarrow \infty$, the $k$th
largest eigenvalue of $S/(n\sigma^2)$ will converge to an upwardly
biased version of $\lambda_{k}+1$ if $\lambda_k$ is greater than
$\sqrt{\alpha}$ \citep{Baik2006, Paul2007}.  This has led several
authors to suggest estimating $\Sigma$ via shrinkage of the
eigenvalues of the sample covariance matrix. In particular, in the
setting that $\sigma^2$ is known, \citet{Donoho2013} propose
estimating all eigenvalues whose sample estimates are smaller than
$\sigma^2(1+\sqrt{\alpha})^2$ by $\sigma^2$, and shrinking the larger
eigenvalues in a way that depends on the particular loss function
being used.  These shrinkage functions are shown to be asymptotically
optimal in the $p/n\rightarrow \alpha$ setting.

Note that covariance estimators of this form are are equivariant with
respect to rotations and scale changes.  The situation should be
different, however, when we are interested in estimating multiple
covariance matrices from distinct but related groups.  Here,
group-level equivariance to rotations is an unreasonable assumption;
both eigenvalue \emph{and} eigenvector shrinkage can play an important
role in improving covariance estimates.  Consider multi-group
covariance estimation based on $K$ matrices, $Y_1, ..., Y_K$, where
$Y_k$ is assumed to be an $n_k$ by $p$ matrix of mean-zero normal
data, typically with $n_k \ll p$.  Then, $Y_k^TY_k = S_k$ has a
(degenerate) Wishart distribution as in Equation \ref{eqn:lik}.  To
improve estimation, we seek estimators of each covariance matrix,
$\hat{\Sigma}_k$, that may depend on data from all groups.  To this
end, we extend the spiked covariance model to grouped data, by
assuming that the anisotropic variability from each group occurs on a
common low dimensional subspace.  Specifically, we posit that the
covariance matrix for each group can be written as
\begin{equation}
\Sigma_k = V\Psi_kV^T + \sigma^2_kI,
\label{eqn:sspsi}
\end{equation}

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/3dplot}
        \caption{Projection in $\mathbb{R}^3$}
        \label{fig:3dplot}
    \end{subfigure}
\quad
 %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter}
        \caption{$Y_kV$}
        \label{fig:2dscatter}
    \end{subfigure}
\quad 
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter-orth}
        \caption{$Y_kV_{\perp}$}
        \label{fig:2dscatterorth}
    \end{subfigure}
    \caption{ Two groups of four-dimensional data (red and blue)
      projected into different subspaces.  a) To visualize $Y_k$ we
      can project the data into $\mathbb{R}^3$.  In this illustration, the
      distributional differences between the groups are confined to a
      two-dimensional shared subspace ($VV^T$, grey plane).  b) The
      data projected onto the two-dimensional shared subspace,
      $Y_kV$, have covariances $\Psi_k$ that differ between
      groups. c) The orthogonal projection, $Y_kV_{\perp}$
      has isotropic covariance, $\sigma_k^2I$, for all groups.  }
\label{fig:shared}
\end{figure}
% For multi-group covariance estimation, we assume that we have $K$
% matrices, $Y_1, ... Y_K$, where $Y_k$ consists of $n_k$ independent
% rows of mean-zero normal data of dimension $p$, typically with
% $n_k \ll p$.

%
\noindent where the columns of $V \in \mathcal{V}_{p, s}$, with
$s \ll p$, determine a subspace shared by all groups.  Throughout this
paper we will denote to the shared subspace as
$VV^T \in \mathcal G_{p,s}$, where $\mathcal G_{p,s}$ is the
Grassmann manifold consisting of all $s$-dimensional linear subspaces
of $\mathbb{R}^p$ \citep{Chikuse2012}.  Although $V$ is only
identifiable up to right rotations, the matrix $VV^T$, which defines
the plane of variation shared by all groups, is identifiable.  Later,
to emphasize the connection to the spiked PCA model
(\ref{eqn:spiked}), we will write $\Psi_k$ in terms of its
eigendecomposition, $\Psi_k = O_k\Lambda_kO_k$, where $O_k$ are
eigenvectors and $\Lambda_k$ are the eigenvalues of $\Psi_k$ (see
Section \ref{sec:bayes}).

@@

In this section we outline an empirical Bayes approach for estimating
a low-dimensional shared subspace and the covariance matrices of the
data projected onto this space. As we discuss in Section
\ref{sec:simulation}, if the spiked covariance model holds for each
group individually, then the shared subspace assumption also holds,
where the shared subspace is simply the span of the group-specific
eigenvectors, $U_1, ..., U_K$. In practice, we can usually identify a
shared subspace of dimension $s \ll p$ that preserves most of the
variation in the data.  Our primary objective is to identify the
``best'' shared subspace of fixed dimension $s < p$.  In Section
\ref{sec:em} we describe an expectation-maximization algorithm for
estimating the maximum marginal likelihood of the shared subspace,
$VV^T$.  This approach is computationally tractable for
high-dimensional datasets.  Given an inferred subspace, we then seek
estimators for the covariance matrices of the data projected onto this
space.  Because seemingly large differences in the point estimates of
covariance matrices across groups may not actually reflect
statistically significant differences, in Section \ref{sec:bayes} we
also describe a Gibbs sampler that can be used to generate estimates
of the projected covariance matrices, $\Psi_k$, and and their
associated uncertainty.  Later, in Section \ref{sec:simulation} we
discuss strategies for inferring an appropriate value for $s$ and
explore how shared subspace models can be used for exploratory data
analysis by visualizing covariance heterogeneity on two or three
dimensional subspaces.

@@

\begin{align}
\nonumber p(S_1, ... S_k | \Sigma_k,n_k) &\propto \prod_{k=1}^K |\Sigma_k|^{-n_k/2}\etr(-\Sigma_k^{-1}S_k/2)  \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-(\sigma_k^2(V\Psi_kV^T +
  I))^{-1}S_k/2) \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-\left[V(\Psi_k +
  I)^{-1}/\sigma_k^2 V^T + (I-VV^T)/\sigma^2_k\right]S_k/2)
  \\
&\propto \prod_{k=1}^K  (\sigma_k^2)^{-n_k(p-s)/2}|M_k|^{-n_k/2}\etr(-\left[VM_k^{-1}V^T + \frac{1}{\sigma^2_k} (I-VV^T)\right]S_k/2) ,
\end{align}
%
\noindent where we define $M_k = \sigma_k^2(\Psi_k + I)$.  The log-likelihood in
$V$ (up to an additive constant) is
%
\begin{align}
\nonumber l(V) &= \sum_k \tr\left(-VM_k^{-1}V^T +
       VV^T/\sigma^2_k\right)S_k/2)\\
&= \frac{1}{2}\sum_k \tr\left(-M_k^{-1}V^T S_kV\right) + \text{tr}\left(VV^T S_k\right)/\sigma_k^2.
\end{align}

@@

We maximize the marginal likelihood of $V$ with an EM algorithm, where
$M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ are considered the
``missing'' parameters.  We assume independent Jeffreys'
prior distributions for both $\sigma_k^2$ and $M_k$.  Jeffreys' prior
for these quantities corresponds to
$p(\sigma_k^2) \propto 1/\sigma_k^2$ and
$p(M_k) \propto |M_k|^{-(s+1)/2}$.  From the likelihood it can easily
be shown that the conditional posterior for $M_k$ is
%
$$p(M_k | V) \propto |M_k|^{(n_k + s + 1)/2}\etr(-(M_k^{-1}V^TS_kV)/2) $$
%
\noindent which is an inverse-Wishart($V^TS_kV$, $n_k$) distribution.  The
conditional posterior distribution of $\sigma_k^2$ is simply
%
$$p\left(\sigma^2 | V\right) \propto (\sigma_k^2)^{-n_k(p-s)/2-1}\etr\left(- (I-VV^T)S_k/[2\sigma_lk^2]\right)  $$
%
\noindent which is an inverse-gamma($n_k(p-s)/2$,
$\text{tr}[(I-VV^T)S_k]/2$) distribution.  Based on these
results, we complete the following steps for each iteration $t$ of
the EM algorithm until convergence:

@@

For the second step (``M-step''), we use a numerical optimization algorithm based on
the Cayley transform to preserve the orthogonality constraints in $V$
\citep{Wen2013}.  Importantly, the complexity of this algorithm is
dominated by the dimension of the shared subspace, not the number of
features.  Thus, our approach is computationally efficient for
relatively small values of $s$, even when $p$ is large.

@@

If $V$ is a basis for a shared subspace, then for each group, most
of the non-isotropic variation in $Y_k$ should be preserved when
projecting the data onto this space.  To characterize the extent to
which this is true for different groups, we propose a simple estimator
for the proportion of ``signal'' variance that lies on a given
subspace.  Specifically, we use the following statistic for the ratio
of the sum of the first $s$ eigenvalues of
$V^T \Sigma_k V$ to the sum of the first $s$ eigenvalues
of $\Sigma_k$:
%
\begin{equation}
\gamma(Y_k: V, \sigma_k^2) = \frac{||Y_kV||_F/n_k}{\underset{\widetilde{V} \in \mathcal{V}_{p, s}}{\text{max}}
  ||Y_k\widetilde{V}||_F/n_k - \sigma_k^2ps/n_k}
\label{eqn:ratio}
\end{equation}
%
\noindent where $||\cdot||_F$ is the Frobenius norm. Our motivation for
this statistic is the fact that if
$\Sigma_k = V\Psi_kV^T + \sigma_k^2I$ then
$\gamma(Y_k: V, \sigma_k^2) \approx 1$.  We establish
this in an asymptotic regime where $s$ is fixed, $p$,
$n_k \rightarrow \infty$ and $p/n_k = \alpha_k$ constant.

@@

As a consequence, the proposed statistic will be close to one for all
groups when $VV^T$ is a shared subspace for the data and smaller if
not.  The metric provides a useful indicator of which groups can be
reasonably compared on a given subspace and which groups cannot.  In
practice, we estimate a shared subspace $\hat{V}$ and the isotropic
variances $\hat{\sigma}_k^2$ using EM and compute the plug-in
estimate $\gamma(Y_k: \hat{V}, \hat{\sigma}_k^2)$.  When this
statistic is small for some groups, it may suggest that the rank $s$
of the inferred subspace needs to be larger to capture the variation
in all groups. We investigate this in Section \ref{sec:simulation}, by
computing the goodness of fit statistic for inferred subspaces of
different dimensions on a single dataset. In Section \ref{sec:app}, we
compute the estimates for subspaces inferred with real biological
data.

@@

The EM algorithm presented in the previous section yields point
estimates for $VV^T$, $\Psi_k$, and $\sigma_k^2$ but does not lead to natural
uncertainty quantification for these estimates.  In this section, we
assume that the subspace $VV^T$ is fixed and known and demonstrate how
we can estimate the posterior distribution for $\Psi_k$.   Note that when
the subspace is known, the posterior distribution of $\Sigma_k$ is
conditionally independent from the other groups, so that we can
independently estimate the conditional posterior distributions for each
group.
